Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations

### 2017-02-27 ParisEM.tex
addition of three boudness assumption

\newcommand{\x}[2]{x_{#1}^{(#2)}}
\newcommand{\w}[2]{w_{#1}^{(#2)}}
\newcommand{\tw}[2]{\tilde{w}_{#1}^{(#2)}}
\newcommand{\p}[2]{\xi_{#1}^{(#2)}}
\newcommand{\tp}[2]{\tilde{\xi}_{#1}^{(#2)}}
\newcommand{\ta}[2]{\tau_{#1}^{(#2)}}
\newcommand{\rmd}{\mathrm{d}}
\newcommand{\eqsp}{\;}
\newcommand{\1}{\mathrm{1}}
\newcommand{\com}[1]{{\color{gray} // #1}}
\newcommand{\mP}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\qk}{q_{k}}
\newcommand{\acom}[1]{\textit{\color{gray} //#1}}
\newcommand{\Oz}{Z}%Letter for the Ozaki approximation
\newcommand{\Jk}{J_{\alpha}^k}%Command for Jacobian of alpha
\newcommand{\mw}{\mathsf{w}}%For bridge realisations
\newcommand{\U}{\mathsf{U}}
\newcommand{\Lo}{\mathsf{L}}
%\newcommand{\qk}{q^{\Delta t_k}_{\theta}}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newcommand{\hQ}{\widehat{Q}}
\newcounter{hypA}
\newenvironment{hypA}{\refstepcounter{hypA}\begin{itemize}
\item[{\bf H\arabic{hypA}}]}{\end{itemize}}

\begin{document}

@@

%Statistical inference for partially observed stochastic differential equations is a challenging task since some elementary quantities, such as transition probabilities, are not available explicitly.
Statistical inference for partially observed stochastic differential equations often requires to solve bayesian filtering and smoothing problems, i.e. the computation of the posterior distributions of sequences of hidden states given observations. The filtering problem refers to the estimation, for each $0\le k \le n$,  of the distributions of the hidden state $X_k$ given the observations $(Y_0,\ldots,Y_k)$. Smoothing stands for the estimation of the distributions of the sequence of states $(X_{k},\ldots,X_{p})$ given observations $(Y_{0},\ldots,Y_{\ell})$ with $0\le k\le p \le \ell \le n$. 
These posterior distributions are crucial to compute maximum likelihood estimators of the unknown model parameters using the observations $(Y_0,\ldots,Y_n)$ only.
% $\theta\mapsto \ell_{\theta}^{n}$ given by
%\begin{equation*}
%\ell_{\theta}^{n}(Y_{0:n}) = \log\left(\int p_{\theta}(x_{0:n},Y_{0:n})\,\rmd x_{0:n}\right)\eqsp,
%\end{equation*}  
%where $p_{\theta}$ is the complete data likelihood:
%\begin{equation*}
%p_{\theta}(x_{0:n},Y_{0:n}) = \chi(x_0)g_{0}(x_0)\prod^{n-1}_{k=0}\qk(x_{k},x_{k+1})g_{k+1}(x_{k+1})\eqsp.
%\end{equation*}
For instance, the E-step of the EM algorithm introduced in \cite{dempster:laird:rubin:1977} %iteratively builds a sequence of parameter estimates following two steps.
%following the two steps:
%\begin{enumerate}
%	\item {\bf E-step}: compute $\theta \mapsto Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log p_{\theta}(X_{0:n},Y_{0:n})\middle|Y_{0:n}\right]$, where $\mathbb{E}_{\theta}\left[\cdot\middle|Y_{0:n}\right]$ is the conditional expectation given $Y_{0:n}$ when the parameter value is $\theta$.
%	\item {\bf M-step}: choose $\theta_{p+1}$ as a maximizer of $\theta \mapsto Q(\theta,\theta_{p})$.
%\end{enumerate}
%step 
boils down to the computation of a conditional expectation of an additive functional of the hidden states given all the observations up to time $n$. 
%\begin{multline*}
%Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log \left(\chi(X_0)g_{0}(X_0)\right)\middle|Y_{0:n}\right] \\
%+ \sum_{k=0}^{n-1}\mathbb{E}_{\theta_p}\left[\log \left(\qk(X_{k},X_{k+1})g_{k+1}(X_{k+1})\right)\middle|Y_{0:n}\right] \eqsp.
%\end{multline*}
Similarly, by Fisher's identity, recursive maximum likelihood estimates may be computed using the gradient of the loglikelihood which can be written as the conditional expectation of an additive functional of the hidden states.
 See \cite[Chapter $10$ and $11$]{cappe:moulines:ryden:2005}, \cite{kantas:doucet:signh:2015,doucet:poyiadjis:singh:2011,lecorff:fort:2013a,lecorff:fort:2013b}
for further references on the use of these smoothed expectations of additive functionals  applied to maximum likelihood parameter inference in latent data models.

@@

 The only solution to solve the smoothing problem for partially observed SDE using SMC methods has been proposed in \cite{olsson:strojby:2011} and extends the fixed-lag smoother of \cite{olsson:cappe:douc:moulines:2008}. 
 Using forgetting properties of the hidden chain, the algorithm improves the performance of \cite{fearnhead:papaspiliopoulos:roberts:2008} to approximate smoothing distributions but at the cost of a bias that does not vanish as the number of particles grows to infinity.
In the case of discrete time state space models, approximations of the smoothing distributions may also be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}. 
Both algorithms require first a forward pass which produces a set of particles and weights approximating the sequence of filtering distributions up to time $n$. Then, a backward pass is performed to compute new weights (FFBS) or sample trajectories (FFBSi) in order to approximate the smoothing distributions. 
%In the case of partially observed SDE, this forward pass may be performed with the SMC steps combined with the GPE of \cite{fearnhead:papaspiliopoulos:roberts:2008}. 
%Then, the backward pass of the FFBS algorithm keeps all the particles sampled during the forward pass and computes new weights to take into account the information brought by all the observations from time $n$ to time $0$. Instead of computing new weights, the FFBSi  algorithm samples independently backward in time particle trajectories using the particles and weights produced by the forward pass. 
Recently, \cite{olsson:westerborn:2016} proposed a new SMC algorithm, the particle-based rapid incremental smoother (PaRIS), to approximate on-the-fly (i.e. using the observations as they are received) smoothed expectations of additive functionals. 
Unlike the FFBS algorithm, the complexity of this algorithm grows only linearly with the number of particles $N$ and contrary to the FFBSi algorithm, no backward pass is required.

@@

Therefore, to obtain a generalized random version of PaRIS, we only need to be able to sample from weight $\Lambda_k^N(i,\ell)$ in the case of SDE based HMM. If $\widehat{q}_k$ may be upper bounded almost surely by a constant $\hat{\sigma}_+$, this task is performed using the accept-reject mechanism given in Algorithm~\ref{alg:AR:unknownq}. Lemma~\ref{lem:AR:unbiased} states that replacing  $\qk$ by its unbiased approximation is a valid alternative to step~\eqref{it:PaRIS:sampleindex}.\\

@@

Let $0\le k\le n-1$.  For all $1\le i \le N$ and all $1\le \ell \le \widetilde{N}$, the conditional probability distribution given $\mathcal{G}_{k+1}^N$ of the random variable $J_{k}^{i,\ell}$ produced by Algorithm~\ref{alg:AR:unknownq} is  $\Lambda_{k-1}^N(i,\cdot)$. 
Note that the assumption that for all $1\le j\le N$, $\sup_y \hat{q}_k(\xi_k^j,y,\zeta_k)< \hat{\sigma}_+^k$ almost surely is actually sufficient to apply Algorithm \ref{alg:AR:unknownq}. This improvement keeps the linear complexity property of PaRIS algorithm and Algorithm \ref{alg:Ozaki:PaRIS}, and allows to increase significantly the acceptance rate of Algorithm \ref{alg:AR:unknownq}. %Moreover, the AR algorithm still stands if one can define 
\begin{algorithm}[H]
\caption{Random weight accept-reject backward sampling}

@@

\FORALL{$i \in 1,\dots, N$}
\FORALL{$\ell \in 1,\dots, \widetilde N$} 
\STATE \textbf{\sc Sampling Step} \\
Sample independently $

@@

\ENDFOR
\ENDFOR

@@

\label{alg:AR:unknownq}
\end{algorithm}

@@

On the r.h.s. of \eqref{eq:GPE1}, the product over $\kappa$ elements is bounded by 1, therefore, a sufficient condition to obtain an upper bound  $\hat{q}_k$ is that the function:

@@

is upper bounded almost surely by $\hat{\sigma}_+^k$.
As said above, the GRand PaRIS algorithm only requires a bound $\hat{\sigma}_+^k$ for the function $\rho_{\Delta_k}$ where $x \in \{\xi_{k}^1,\dots,\xi_{1}^N\}$. In order to improve the acceptance rate (at a cost of a quadratic complexity in $N$), we can always compute, for all $1\le i\le N$, the bound 
\hat{\sigma}_{+}^{k,i}= \underset{1 \le j\le N}{\text{sup}}\rho_{\Delta_k}(\xi_{k-1}^j,\xi_k^i)\eqsp, \label{eq:quadr:bound}
as long as $\Lo_\mw$ is bounded almost surely. This might increase significantly the acceptance rate of the algorithm, and therefore reduce the number of drawings of the random variable $\zeta_k$, which has a much higher cost than the computation of $\rho$, as it requires simulation of Brownian Bridges. It is worth noting here that this latter option can also avoid numerical optimization if no analytical version of $\hat{\sigma}_+^k$ is available. In the example of the SINE model presented below, for $N=400$ particles, the algorithm with linear complexity  has a 40\% higher computation time than the algorithm with a quadratic complexity. Actually, this condition is always satisfied for models in the domains $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, i.e. domains for which the exact algorithms EA1 and EA2 can be used.

%\subsection*{Note on the complexity and practical implementation}
%As said above, as long as one can define a $\sigma_+^k$ such that
%$$ \forall\eqsp 0\leq k\leq n-1,\eqsp 1\leq i\leq N,\eqsp y\in\mathbb{R}^d,\eqsp \hat{q}_k(\xi_k^i,y,\zeta_k)<\hat{\sigma}_+^k \text{ a.s.,}$$ then the resulting GRand Paris algorithm has a linear complexity in $N$.\\
%However, it is worth noting here that for a reasonable amount of particles, it might be more efficient in terms of computation time to compute, for all $i$, the bound given by \eqref{eq:quadr:bound}. Indeed, this bound requires, at each time step $k$, $N^2$ computation of the function $\rho$ given above. However, this might increase significantly the acceptance rate of the AR algorithm, and therefore reduce the number of drawings for the random variable $\zeta_k$, which has a much higher cost than the computation of $\rho$, as it requires simulation of Brownian Bridges. It is worth noting here that this latter option can also avoid numerical optimization if no analytical version of $\hat{\sigma}_+^k$ is available.\\
%In the example of the SINE model presented below, for $N=400$ particles, the linear complexity algorithm has a 40\% higher computation time than the algorithm with a quadratic complexity. In practice, in addition to $n$ and $N$, the computation time will increase with
%\begin{itemize}
%\item $\tilde{N}$ the number of retrospective sampling;
%\item $M$ the number of Monte Carlo samples to obtain $\hat{q}$;
%\item $\Delta$ the acquisition time step;
%\item The expected value of $\U_\mw-\Lo_\mw$.
%\end{itemize}

@@

\section{Numerical experiments}
\label{sec:exp}
%This section investigates the performance of the proposed algorithm. The different tuning parameters are:
%\begin{enumerate}[i)]
%\item the number of particles $N$ and the precision parameter $\widetilde{N}$;
%\item the proposal transition density $p_k$ and the adjustment multiplier $\vartheta_k$ used to propagate and weight particles;
%\item the discrete distribution $\mu$ used to define $\widetilde{q}_{\theta}^{\Delta t_k}$;
%\item the number of samples to compute $\widetilde{q}_{\theta}^{\Delta t_k}$.
%\end{enumerate}
%The performance of the algorithm is investigated on two known models based on diffusion process.
This section investigates the performance of the proposed algorithm with the sine and log-growth models. In both cases, the proposal distribution $p_k$ is chosen as the following approximation of the optimal filter (or the fully adapted particle filter in the terminology of \cite{pitt:shephard:1999}): 
$$p_k(x_{k-1},x_k)\propto \tilde{q}_k(x_{k-1}, x_k)g_k(x_{k})\eqsp,$$
where $\tilde{q}_k(x_{k-1},x_k)$ is the p.d.f. of Gaussian distibution with mean $\alpha(x_{k-1})\Delta_k$ and variance $\Delta_kI_d$, i.e. the Euler approximation of equation \eqref{eq:target:sde}. As the observation model is linear and Gaussian, the proposal distribution is therefore Gaussian with explicit mean and variance.

@@

%The goal is to simulate realisations of  a discrete random variable $X$ having the following target distribution:
%\begin{equation}
%\mP(X=\ell)= p_X(\ell)=\frac{\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})}{\sum_{\ell=1}^N\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})},~~\ell\in \{1,\dots,N\}
%\end{equation}
%for some $\xi_{k+1}$ that has no importance in this result. We assume $\omega_k$ are normalized weights, i.e. $\sum_\ell \omega_k^l=1$. The candidate will therefore be sampled from $Y$, a r.v. with the following distribution:
%\begin{equation}
%\mP(Y=\ell)=p_Y(\ell)= \omega_k^\ell,~~\ell\in \{1,\dots,N\}
%\end{equation}
%Unfortunately, the $q_k$ function is unknown. Howevere, we let's suppose we can define an unbiased estimator $\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)$, where $\zeta$ is a random variable having th p.d.f. $p_\zeta$.  We have therefore:
%$$\E_{p_\zeta}\left[\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)\right]=q(\xi_k^\ell,\xi_{k+1})$$
%Moreover, we impose the following property over this estimator, we want that:
%$$\exists \sigma_+\in \mathbb{R}\text{ such that }\forall x,y,~~\hat{q}_k(x,y,\zeta)\leq \sigma_+$$
%To simulate a random variable $X\sim p_X$, we propose the following algorithm:
%\begin{enumerate}
%\item Sample independently $\ell\sim p_Y$, $z\sim p_\zeta$ and $u \sim \mathcal{U}[0,1]$
%\item \textbf{If} 
%$$u \leq \frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},z)}{\sigma_+},$$
%set $x=\ell$. \textbf{Else}, return to 1).
%\end{enumerate}

>>>>

\newcommand{\x}[2]{x_{#1}^{(#2)}}
\newcommand{\w}[2]{w_{#1}^{(#2)}}
\newcommand{\tw}[2]{\tilde{w}_{#1}^{(#2)}}
\newcommand{\p}[2]{\xi_{#1}^{(#2)}}
\newcommand{\tp}[2]{\tilde{\xi}_{#1}^{(#2)}}
\newcommand{\ta}[2]{\tau_{#1}^{(#2)}}
\newcommand{\rmd}{\mathrm{d}}
\newcommand{\eqsp}{\;}
\newcommand{\1}{\mathrm{1}}
\newcommand{\com}[1]{{\color{gray} // #1}}
\newcommand{\mP}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\qk}{q_{k}}
\newcommand{\acom}[1]{\textit{\color{gray} //#1}}
\newcommand{\Oz}{Z}%Letter for the Ozaki approximation
\newcommand{\Jk}{J_{\alpha}^k}%Command for Jacobian of alpha
\newcommand{\mw}{\mathsf{w}}%For bridge realisations
\newcommand{\U}{\mathsf{U}}
\newcommand{\Lo}{\mathsf{L}}
%\newcommand{\qk}{q^{\Delta t_k}_{\theta}}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newcommand{\hQ}{\widehat{Q}}
\newcounter{hypA}
\newenvironment{hypA}{\refstepcounter{hypA}\begin{itemize}
\item[{\bf H\arabic{hypA}}]}{\end{itemize}}
%%ENvironmment assumption
\newcounter{defcounter}
\setcounter{defcounter}{0}
\newenvironment{assumpt}{%
\addtocounter{equation}{-1}
\refstepcounter{defcounter}
\renewcommand\theequation{\textbf{A}\thedefcounter}
\begin{equation}}
{\end{equation}}
\begin{document}

@@

%Statistical inference for partially observed stochastic differential equations is a challenging task since some elementary quantities, such as transition probabilities, are not available explicitly.
Statistical inference for partially observed stochastic differential equations often requires to solve bayesian filtering and smoothing problems, i.e.\ the computation of the posterior distributions of sequences of hidden states given observations. The filtering problem refers to the estimation, for each $0\le k \le n$,  of the distributions of the hidden state $X_k$ given the observations $(Y_0,\ldots,Y_k)$. Smoothing stands for the estimation of the distributions of the sequence of states $(X_{k},\ldots,X_{p})$ given observations $(Y_{0},\ldots,Y_{\ell})$ with $0\le k\le p \le \ell \le n$. 
These posterior distributions are crucial to compute maximum likelihood estimators of the unknown model parameters using the observations $(Y_0,\ldots,Y_n)$ only.
% $\theta\mapsto \ell_{\theta}^{n}$ given by
%\begin{equation*}
%\ell_{\theta}^{n}(Y_{0:n}) = \log\left(\int p_{\theta}(x_{0:n},Y_{0:n})\,\rmd x_{0:n}\right)\eqsp,
%\end{equation*}  
%where $p_{\theta}$ is the complete data likelihood:
%\begin{equation*}
%p_{\theta}(x_{0:n},Y_{0:n}) = \chi(x_0)g_{0}(x_0)\prod^{n-1}_{k=0}\qk(x_{k},x_{k+1})g_{k+1}(x_{k+1})\eqsp.
%\end{equation*}
For instance, the E-step of the EM algorithm introduced in \cite{dempster:laird:rubin:1977} %iteratively builds a sequence of parameter estimates following two steps.
%following the two steps:
%\begin{enumerate}
%	\item {\bf E-step}: compute $\theta \mapsto Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log p_{\theta}(X_{0:n},Y_{0:n})\middle|Y_{0:n}\right]$, where $\mathbb{E}_{\theta}\left[\cdot\middle|Y_{0:n}\right]$ is the conditional expectation given $Y_{0:n}$ when the parameter value is $\theta$.
%	\item {\bf M-step}: choose $\theta_{p+1}$ as a maximizer of $\theta \mapsto Q(\theta,\theta_{p})$.
%\end{enumerate}
%step 
boils down to the computation of a conditional expectation of an additive functional of the hidden states given all the observations up to time $n$. 
%\begin{multline*}
%Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log \left(\chi(X_0)g_{0}(X_0)\right)\middle|Y_{0:n}\right] \\
%+ \sum_{k=0}^{n-1}\mathbb{E}_{\theta_p}\left[\log \left(\qk(X_{k},X_{k+1})g_{k+1}(X_{k+1})\right)\middle|Y_{0:n}\right] \eqsp.
%\end{multline*}
Similarly, by Fisher's identity, recursive maximum likelihood estimates may be computed using the gradient of the loglikelihood which can be written as the conditional expectation of an additive functional of the hidden states.
 See \cite[Chapter $10$ and $11$]{cappe:moulines:ryden:2005}, \cite{kantas:doucet:signh:2015,doucet:poyiadjis:singh:2011,lecorff:fort:2013a,lecorff:fort:2013b}
for further references on the use of these smoothed expectations of additive functionals  applied to maximum likelihood parameter inference in latent data models.

@@

 The only solution to solve the smoothing problem for partially observed SDE using SMC methods has been proposed in \cite{olsson:strojby:2011} and extends the fixed-lag smoother of \cite{olsson:cappe:douc:moulines:2008}. 
 Using forgetting properties of the hidden chain, the algorithm improves the performance of \cite{fearnhead:papaspiliopoulos:roberts:2008} to approximate smoothing distributions but at the cost of a bias that does not vanish as the number of particles grows to infinity.
In the case of discrete time state space models, approximations of the smoothing distributions may also be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}. 
Both algorithms require first a forward pass which produces a set of particles and weights approximating the sequence of filtering distributions up to time $n$. Then, a backward pass is performed to compute new weights (FFBS) or sample trajectories (FFBSi) in order to approximate the smoothing distributions. 
%In the case of partially observed SDE, this forward pass may be performed with the SMC steps combined with the GPE of \cite{fearnhead:papaspiliopoulos:roberts:2008}. 
%Then, the backward pass of the FFBS algorithm keeps all the particles sampled during the forward pass and computes new weights to take into account the information brought by all the observations from time $n$ to time $0$. Instead of computing new weights, the FFBSi  algorithm samples independently backward in time particle trajectories using the particles and weights produced by the forward pass. 
Recently, \cite{olsson:westerborn:2016} proposed a new SMC algorithm, the particle-based rapid incremental smoother (PaRIS), to approximate on-the-fly (i.e.\ using the observations as they are received) smoothed expectations of additive functionals. 
Unlike the FFBS algorithm, the complexity of this algorithm grows only linearly with the number of particles $N$ and contrary to the FFBSi algorithm, no backward pass is required.

@@

Therefore, to obtain a generalized random version of PaRIS, we only need to be able to sample from weight $\Lambda_k^N(i,\ell)$ in the case of SDE based HMM.\\
We need to define the following assumption:\\
\begin{assumpt}
\label{assumpt:global}
\forall~0\leq k\leq n,~\exists \hat{\sigma}_+ \text{ such that }\forall (x,y) \in\mathbb{R}^d\times  \mathbb{R}^d,~~\widehat{q}_k(x,y,\zeta_k)\leq \hat{\sigma_+}
\end{assumpt}
Then, the following lemma provides the missing step of a Generalized Random PaRIS (GRand Paris)

@@

Let $0\le k\le n-1$ and $1\le i \le N$, and suppose $A_1$ holds may be upper bounded almost surely by a constant $\hat{\sigma}_+$. Let's define the random variable $J_k^i$ as follow:

@@

\STATE \textsc{Sampling step:} Sample independently $

@@

@@

Then, the conditional probability distribution given $\mathcal{G}_{k+1}^N$ of $J_{k}^{i,\ell}$ is $\Lambda_{k}^N(i,\cdot)$. 
Note that lemma \ref{lem:AR:unbiased} still holds if assumption \eqref{assumpt:global} is relaxed and replaced by one of the two following assumptions:
\begin{assumpt}\label{assumpt:ysupport}
\forall~0\leq k\leq n,~\exists \hat{\sigma}_+ \text{ such that }\forall (j,y) \in [[1,N]]\times  \mathbb{R}^d,~~\widehat{q}_k(\xi^j_k,y,\zeta_k)\leq \hat{\sigma}_+
\end{assumpt}
\vspace{-\baselineskip}
\begin{assumpt}\label{assumpt:local}
\forall~0\leq k\leq n,~\exists \hat{\sigma}_+ \text{ such that }\forall j \in [[1,N]],~~\widehat{q}_k(\xi^j_k,\xi^j_{k+1},\zeta_k)\leq \hat{\sigma}_+
\end{assumpt}
It is worth noting that under assumptions \eqref{assumpt:global} or \eqref{assumpt:ysupport}, the linear complexity property of PaRIS algorithm still holds, whereas if only assumption \eqref{assumpt:local} holds, the resulting algorithm has a quadratic complexity.
%Moreover, the AR algorithm still stands if one can define 

@@

On the r.h.s. of \eqref{eq:GPE1}, the product over $\kappa$ elements is bounded by 1, therefore, a sufficient condition to satisfy of the assumptions \eqref{assumpt:global}-\eqref{assumpt:local} is that the function:

@@

is upper bounded almost surely by $\hat{\sigma}_+$.
In particular, if $Lo_\mw$ is bounded almost surely, \eqref{eq:rho:func} always satisfies assumption \eqref{assumpt:local}, therefore, algorithm \eqref{alg:Ozaki:PaRIS} can be perfomed, in the worse case with a quadratic complexity in $N$. Therefore, this condition is always satisfied for models in the domains $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, i.e.\ domains for which the exact algorithms EA1 and EA2 can be used.\\ 
In that case, and when \eqref{assumpt:global} or \eqref{assumpt:ysupport} holds, it can be nonetheless of practical interest to choose the $\hat{\sigma}_+$ corresponding to \eqref{assumpt:global}. Indeed, this might increase significantly the acceptance rate of the algorithm, and therefore reduce the number of drawings of the random variable $\zeta_k$, which has a much higher cost than the computation of $\rho$, as it requires simulation of Brownian Bridges. 
Moreover, this latter option can also avoid numerical optimization if no analytical version of $\hat{\sigma}_+^k$ is available. In practice, we found this option more efficient in terms of computation time when $N$ has moderate values. 
%\subsection*{Note on the complexity and practical implementation}
%As said above, as long as one can define a $\sigma_+^k$ such that
%$$ \forall\eqsp 0\leq k\leq n-1,\eqsp 1\leq i\leq N,\eqsp y\in\mathbb{R}^d,\eqsp \hat{q}_k(\xi_k^i,y,\zeta_k)<\hat{\sigma}_+^k \text{ a.s.,}$$ then the resulting GRand Paris algorithm has a linear complexity in $N$.\\
%However, it is worth noting here that for a reasonable amount of particles, it might be more efficient in terms of computation time to compute, for all $i$, the bound given by \eqref{eq:quadr:bound}. Indeed, this bound requires, at each time step $k$, $N^2$ computation of the function $\rho$ given above. However, this might increase significantly the acceptance rate of the AR algorithm, and therefore reduce the number of drawings for the random variable $\zeta_k$, which has a much higher cost than the computation of $\rho$, as it requires simulation of Brownian Bridges. It is worth noting here that this latter option can also avoid numerical optimization if no analytical version of $\hat{\sigma}_+^k$ is available.\\
%In the example of the SINE model presented below, for $N=400$ particles, the linear complexity algorithm has a 40\% higher computation time than the algorithm with a quadratic complexity. In practice, in addition to $n$ and $N$, the computation time will increase with
%\begin{itemize}
%\item $\tilde{N}$ the number of retrospective sampling;
%\item $M$ the number of Monte Carlo samples to obtain $\hat{q}$;
%\item $\Delta$ the acquisition time step;
%\item The expected value of $\U_\mw-\Lo_\mw$.
%\end{itemize}

@@

\section{Numerical experiments}
\label{sec:exp}
%This section investigates the performance of the proposed algorithm. The different tuning parameters are:
%\begin{enumerate}[i)]
%\item the number of particles $N$ and the precision parameter $\widetilde{N}$;
%\item the proposal transition density $p_k$ and the adjustment multiplier $\vartheta_k$ used to propagate and weight particles;
%\item the discrete distribution $\mu$ used to define $\widetilde{q}_{\theta}^{\Delta t_k}$;
%\item the number of samples to compute $\widetilde{q}_{\theta}^{\Delta t_k}$.
%\end{enumerate}
%The performance of the algorithm is investigated on two known models based on diffusion process.
This section investigates the performance of the proposed algorithm with the sine and log-growth models. In both cases, the proposal distribution $p_k$ is chosen as the following approximation of the optimal filter (or the fully adapted particle filter in the terminology of \cite{pitt:shephard:1999}): 
$$p_k(x_{k-1},x_k)\propto \tilde{q}_k(x_{k-1}, x_k)g_k(x_{k})\eqsp,$$
where $\tilde{q}_k(x_{k-1},x_k)$ is the p.d.f. of Gaussian distibution with mean $\alpha(x_{k-1})\Delta_k$ and variance $\Delta_kI_d$, i.e.\ the Euler approximation of equation \eqref{eq:target:sde}. As the observation model is linear and Gaussian, the proposal distribution is therefore Gaussian with explicit mean and variance.

@@

%The goal is to simulate realisations of  a discrete random variable $X$ having the following target distribution:
%\begin{equation}
%\mP(X=\ell)= p_X(\ell)=\frac{\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})}{\sum_{\ell=1}^N\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})},~~\ell\in \{1,\dots,N\}
%\end{equation}
%for some $\xi_{k+1}$ that has no importance in this result. We assume $\omega_k$ are normalized weights, i.e.\ $\sum_\ell \omega_k^l=1$. The candidate will therefore be sampled from $Y$, a r.v. with the following distribution:
%\begin{equation}
%\mP(Y=\ell)=p_Y(\ell)= \omega_k^\ell,~~\ell\in \{1,\dots,N\}
%\end{equation}
%Unfortunately, the $q_k$ function is unknown. Howevere, we let's suppose we can define an unbiased estimator $\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)$, where $\zeta$ is a random variable having th p.d.f. $p_\zeta$.  We have therefore:
%$$\E_{p_\zeta}\left[\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)\right]=q(\xi_k^\ell,\xi_{k+1})$$
%Moreover, we impose the following property over this estimator, we want that:
%$$\exists \sigma_+\in \mathbb{R}\text{ such that }\forall x,y,~~\hat{q}_k(x,y,\zeta)\leq \sigma_+$$
%To simulate a random variable $X\sim p_X$, we propose the following algorithm:
%\begin{enumerate}
%\item Sample independently $\ell\sim p_Y$, $z\sim p_\zeta$ and $u \sim \mathcal{U}[0,1]$
%\item \textbf{If} 
%$$u \leq \frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},z)}{\sigma_+},$$
%set $x=\ell$. \textbf{Else}, return to 1).
%\end{enumerate}
